Hybrid nonlinear system identification using neural networks

### 2020-11-23 systemID_NNs.tex
edit.

\documentclass{article}

%% The graphicx and color packages should give you all the color,
%% figure inclusion, and text/figure resizing capabilities you need.
\usepackage{graphicx}
\usepackage[dvipsnames]{xcolor}
\usepackage{tikz,tikzsettings,bm}
\usepackage{almostfull}

@@

\title{Data-based completion of grey-box models with neural networks}
\author{Pratyush Kumar}
\date{\today}

@@

\section{Process modeling}
Consider a nonlinear plant evolving in continuous time as
\begin{align*}
  \dot x_p &= f_p(x_p, u, w) \\
  y &= h_p(x_p) + v
\end{align*}
in which $x_p \in \bbR^{n_p}$ is the plant state, 
$u \in \bbU$ is the manipulated control input,
and $y \in \bbR^p$ is the measurement.

@@

in which $x_g \in \bbR^{n_g}$ are the grey-box states, 
which are some subset 
of the plant states (denote $x_g = Gx_p$) chosen to 
be modeled. The dynamics of these grey-box states 
potentially have some missing terms 
in their ODEs.
For observable nonlinear systems, 
we can construct the original plant state
using a history of measurements and control inputs.
Denote this history as 
\begin{align*}
  \mathbf{y}_{k-N_p:k-1} &= [y(k-\Delta)', ..., y(k-N_p\Delta)']' \\
  \mathbf{u}_{k-N_p:k-1} &= [u(k-\Delta)', 
                               ..., u(k-N_p\Delta)']'
\end{align*}

@@

We augment the grey-box model using 
neural networks as follows -
\begin{align*}
\dot x_g &= f_g(x_g, u) + 
f_N(h_g(x_g), z, u)\\
y &= h_g(x_g) 
\end{align*}
in which $f_N(\cdot)$ is a feedforward neural network
and $z(k) = [\mathbf{y}'_{k-N_p:k-1}, \mathbf{u}'_{k-N_p:k-1}]'$.

Alternative modeling choices are:
\begin{itemize}
	\item Black-box: $y^+ = f_N(y, z, u)$, (purely black-box)
	\item Hybrid 1: $\dot x_g = f_N(x_g, z, u), \ y = h_g(x_g)$, 
  (states have meaning)
  \item Hybrid 2: $\dot x_g = f_g(x_g, u), \ y = h_g(x_g) + h_N(z)$, 
	(model residual)
	\item Hybrid 3: $\dot x_g = f_g(x_g, u, p(h_g(x_g), z)), \ y = h_g(x_g)$, (use NNs to learn some parameters)
\end{itemize}

@@

\subsection{Conversion to discrete time and training}
We note that the network $f_N(\cdot)$ is an addition 
to the grey-box dynamics in continuous time. 
The Runge-Kutta-4 (RK4) method 
is used for the discretization as follows

\begin{align*}
  k_1 &= f_g(x_g, u) + 
  f_N(h_g(x_g), \mathbf{y}_{k-N_p:k-1}, \mathbf{u}_{k+1-N_p:k}) \\
  x_g & := x_g + (\Delta/2)k_1 \\
  k_2 &= f_g(x_g, u) + 
  f_N(h_g(x_g), \mathbf{y}_{k+0.5-N_p:k-0.5}, \mathbf{u}_{k+1-N_p:k})\\
  x_g & := x_g + (\Delta/2)k_2 \\
  k_3 &= f_g(x_g, u) + 
  f_N(h_g(x_g), \mathbf{y}_{k+0.5-N_p:k-0.5}, \mathbf{u}_{k+1-N_p:k})\\
  x_g & := x_g + \Delta k_3 \\
  k_4 &= f_g(x_g, u) + 
  f_N(h_g(x_g), \mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k+1-N_p:k})\\
  x_g(k+\Delta) &= x_g(k) + (\Delta/6)(k_1 + 2k_2 + 2k_3 + k_4) \\
  \hat{y}(k+\Delta) &= h_g(x_g(k+\Delta)) 
\end{align*}
in which the sequence $\mathbf{y}_{k+0.5-N_p:k-0.5}$ is obtained by 
linear interpolation.

@@

\subsection{Linear plant}

Consider the following plant, model,
and the expected function to be approximated 
with a neural network.

\textbf{Plant --}
\begin{align*}
  \dfrac{dC_A}{dt} &= \dfrac{C_{Af} - C_A}{\tau} - k_1C_A\\
  \dfrac{dC_B}{dt} &= k_1C_A - k_2C_B + k_3C_C- \dfrac{C_B}{\tau}\\
  \dfrac{dC_C}{dt} &= k_2C_B - k_3C_C - \dfrac{C_C}{\tau}
\end{align*}
These ODEs represent the dynamics of the reaction system
$A \rightarrow B \rightleftharpoons C$ in a CSTR with constant 
height and temperature. The control input is $u = C_{Af}$,
the states are $x = [C_A, C_B, C_C]'$, 
and the measurements are $y = [C_A, C_B]'$. 
Assume the following 
parameters for this model, $k_1 = 1 \ \textnormal{m}^3/\textnormal{min}, 
\ k_2 = 0.3 \ \textnormal{m}^3/\textnormal{min},
\ k_3 = 0.2 \ \textnormal{m}^3/\textnormal{min}$, 
and $\tau = 5 \ \textnormal{min}$.
The plant model 
is linear ($dx/dt = Ax + Bu$) and 
observable with the choices of 
the parameters and measurements.

\textbf{Model --}
\begin{align*}
  \dfrac{dC_A}{dt} &= \dfrac{C_{Af} - C_A}{\tau} - k_1C_A\\
  \dfrac{dC_B}{dt} &= k_1C_A - \dfrac{C_B}{\tau}
\end{align*}
i.e, we assume that we don't know about the second reaction while 
performing grey-box modeling.

The current plant state $x_p(k)$ can be constructed using the history of 
past measurements and inputs as 
\begin{equation*}
  x_p(k) = A^{N_p-1}_d
  \mathcal{O}^{\dagger}(\mathbf{y}_{k+1-N_p:k} - 
                        \mathcal{B}\mathbf{u}_{k+1-N_p:k-1}) + \\ 
      \mathcal{B}_{nN_p-n+1:nN_p}\mathbf{u}_{k+1-N_p:k-1}
\end{equation*}

in which $A_d$ is the plant state transition matrix in discrete-time, 
$\mathcal{O}$ is the observability matrix, 
and $\mathcal{B}$ is a combination of the 
discrete time matrices ($A_d$ and $B_d$) such that it 
denotes the response of a vector of inputs to states, 
and $\mathcal{B}_{nN_p-n+1:nN_p}$ denotes the last $n$
rows of $\mathcal{B}$.

Since the measurements are a simple linear function 
of the grey-box and plant states, we use only one 
neural network to 
augment the grey-box dynamics. 
The expected function 
to be approximated with the network is
\begin{align*}
  f_N(\mathbf{y}_{k-N_p:k}, \mathbf{u}_{k+1-N_p:k-1}) \approx 
  \begin{bmatrix}
  0 & 0 & 0 \\  
  0 & -k_2 & k_3 
  \end{bmatrix}x_p(k)
\end{align*}

In this example, the network just approximates a linear function.
The Figure \ref{fig:validation_linear} 
below shows validation performances of the grey-box 
and augmented hybrid models. The structure 
of the network is restricted to just a matrix and
the augmented model was trained on 240 samples (4 hours).

THe Figures \ref{fig:cost_nonlinear_5hours} and 
\ref{fig:cost_nonlinear_5hours}

\begin{figure}[!h]
  \centering
  \includegraphics[page=2, width=\textwidth]{tworeac_plots_lin.pdf}
  \caption{Model validation for the linear plant}
  \label{fig:validation_linear}
\end{figure}

We notice that the augmented model accurately predicts the plant 
behavior. The steady cost profile (Cost = $c_1C_{Af} - c_2C_B$)
is shown in Figure \ref{fig:cost_linear}.

\begin{figure}[!h]
  \centering
  \includegraphics[page=3, width=\textwidth]{tworeac_plots_lin.pdf}
  \caption{Steady state cost curve for 
           the linear plant, grey-box, and hybrid model.}
  \label{fig:cost_linear}
\end{figure}

\subsection{Nonlinear plant}
The plant is made nonlinear by changing the side reaction 
to a third order reaction ($A \rightarrow B $ and  $3B \rightleftharpoons C$),
and is now simulated with the following ODEs
\begin{align*}
  \dfrac{dC_A}{dt} &= \dfrac{C_{Af} - C_A}{\tau} - k_1C_A\\
  \dfrac{dC_B}{dt} &= k_1C_A - 3k_2C^3_B + 3k_3C_C- \dfrac{C_B}{\tau}\\
  \dfrac{dC_C}{dt} &= k_2C^3_B - k_3C_C - \dfrac{C_C}{\tau}
\end{align*}

>>>>

\documentclass{article}
\usepackage[total={6in, 8in}]{geometry}
%% The graphicx and color packages should give you all the color,
%% figure inclusion, and text/figure resizing capabilities you need.
\usepackage{graphicx}
\usepackage[dvipsnames]{xcolor}
\usepackage{tikz,tikzsettings,bm}
\usepackage{almostfull}

@@

\title{Hybrid nonlinear system identification using neural networks}
\author{Pratyush Kumar}
\date{\today}

@@

\section{Process modeling}
Consider a nonlinear plant evolving in continuous time as
\begin{align*}
  \dot{x}_p &= f_p(x_p, u, w) \\
  y &= h_p(x_p) + v
\end{align*}
in which $x_p \in \bbR^{n_p}$ is the plant state, 
$u \in \bbU$ is the manipulated control input,
$y \in \bbR^p$ is the measurement, $w$ is the process noise, 
and $v$ is the measurement noise.

@@

in which $x_g \in \bbR^{n_g}$ are the grey-box states, 
which are some subset 
of the plant states (denote $x_g = Gx_p$) chosen to 
be modeled. The dynamics of these grey-box states 
potentially have some missing terms 
in their ODEs.
For observable nonlinear systems, 
we can construct the original plant state
using a history of measurements and control inputs.
Denote this history as 
\begin{align*}
  \mathbf{y}_{k-N_p:k-1} &= [y(k-N_p\Delta)', ... \ , y(k-\Delta)']' \\
  \mathbf{u}_{k-N_p:k-1} &= [u(k-N_p\Delta)', 
                               ... \ , u(k-\Delta)']'
\end{align*}

@@

We augment the grey-box model using
neural networks as follows -
\begin{align} \label{eq:hybrid_cont_time}
\dot x_g &= f_g(x_g, u) + 
f_N(x_g, z, u) \notag \\
y &= h_g(x_g) 
\end{align}
in which $f_N(\cdot)$ is a feedforward neural network
and $z(k) = [\mathbf{y}'_{k-N_p:k-1}, \mathbf{u}'_{k-N_p:k-1}]'$.
This model is converted to discrete-time using the 
Runge-Kutta 4 method, and the vector of past controls
and measurements ($z$) is also viewed as a state. The discrete-time 
model is represented as 
\begin{align*}
  x_g^+ &= f_h(x_g, z, u) \\
  z^+ &= f_z(z, x_g, u)\\
  y &= h_g(x_g) 
\end{align*}
in which, $f_h(\cdot)$ is a obtained by discretization
of \eqref{eq:hybrid_cont_time} using the Runge-Kutta 4 method, 
and $f_z$ is known analytically.

A few alternative modeling choices are:
\begin{itemize}
	\item Black-box: $y^+ = f_N(y, z, u)$, (purely black-box)
  \item Hybrid 1: $x_g^+ = f_N(x_g, z, u), 
                   \ z^+ = f_z(z, x_g, u), \ y = h_g(x_g)$, 
  (states have meaning)
  \item Hybrid 2: $x_g^+ = f_g(x_g, u), 
                   \ z^+ = f_z(z, x_g, u), \ y = h_g(x_g) + h_N(z)$, 
	(model residual)
  \item Hybrid 3: $\dot x_g = f_g(x_g, u, p(h_g(x_g), z)),
  \ z^+ = f_z(z, x_g, u), \ y = h_g(x_g)$, (use NNs to learn some parameters)
\end{itemize}

@@

\subsection{Conversion to discrete time and training}
We note that the network $f_N(\cdot)$ is an addition 
to the grey-box dynamics in continuous time in \eqref{eq:hybrid_cont_time}. 
The Runge-Kutta-4 (RK4) method 
is used for the discretization as follows

\begin{align*}
  k_1 &= f_g(x_g, u) + 
  f_N(x_g, \mathbf{y}_{k-N_p:k-1}, \mathbf{u}_{k-N_p:k-1}, u) \\
  x_g & := x_g + (\Delta/2)k_1 \\
  k_2 &= f_g(x_g, u) + 
  f_N(x_g, \mathbf{y}_{k+0.5-N_p:k-0.5}, \mathbf{u}_{k-N_p:k-1}, u)\\
  x_g & := x_g + (\Delta/2)k_2 \\
  k_3 &= f_g(x_g, u) + 
  f_N(x_g, \mathbf{y}_{k+0.5-N_p:k-0.5}, \mathbf{u}_{k-N_p:k-1}, u)\\
  x_g & := x_g + \Delta k_3 \\
  k_4 &= f_g(x_g, u) + 
  f_N(x_g, \mathbf{y}_{k+1-N_p:k}, \mathbf{u}_{k-N_p:k-1}, u)\\
  x_g(k+\Delta) &= x_g(k) + (\Delta/6)(k_1 + 2k_2 + 2k_3 + k_4) \\
  \hat{y}(k+\Delta) &= h_g(x_g(k+\Delta)) 
\end{align*}
in which the sequence $\mathbf{y}_{k+0.5-N_p:k-0.5}$ is obtained by 
linear interpolation.

@@

%\subsection{Linear plant}

%Consider the following plant, model,
%and the expected function to be approximated 
%with a neural network.

%\textbf{Plant --}
%\begin{align*}
%  \dfrac{dC_A}{dt} &= \dfrac{C_{Af} - C_A}{\tau} - k_1C_A\\
%  \dfrac{dC_B}{dt} &= k_1C_A - k_2C_B + k_3C_C- \dfrac{C_B}{\tau}\\
%  \dfrac{dC_C}{dt} &= k_2C_B - k_3C_C - \dfrac{C_C}{\tau}
%\end{align*}
%These ODEs represent the dynamics of the reaction system
%$A \rightarrow B \rightleftharpoons C$ in a CSTR with constant 
%height and temperature. The control input is $u = C_{Af}$,
%the states are $x = [C_A, C_B, C_C]'$, 
%and the measurements are $y = [C_A, C_B]'$. 
%Assume the following 
%parameters for this model, $k_1 = 1 \ \textnormal{m}^3/\textnormal{min}, 
%\ k_2 = 0.3 \ \textnormal{m}^3/\textnormal{min},
%\ k_3 = 0.2 \ \textnormal{m}^3/\textnormal{min}$, 
%and $\tau = 5 \ \textnormal{min}$.
%The plant model 
%is linear ($dx/dt = Ax + Bu$) and 
%observable with the choices of 
%the parameters and measurements.

%\textbf{Model --}
%\begin{align*}
%  \dfrac{dC_A}{dt} &= \dfrac{C_{Af} - C_A}{\tau} - k_1C_A\\
%  \dfrac{dC_B}{dt} &= k_1C_A - \dfrac{C_B}{\tau}
%\end{align*}
%i.e, we assume that we don't know about the second reaction while 
%performing grey-box modeling.

%The current plant state $x_p(k)$ can be constructed using the history of 
%past measurements and inputs as 
%\begin{equation*}
%  x_p(k) = A^{N_p-1}_d
%  \mathcal{O}^{\dagger}(\mathbf{y}_{k+1-N_p:k} - 
%                        \mathcal{B}\mathbf{u}_{k+1-N_p:k-1}) + \\ 
%      \mathcal{B}_{nN_p-n+1:nN_p}\mathbf{u}_{k+1-N_p:k-1}
%\end{equation*}

%in which $A_d$ is the plant state transition matrix in discrete-time, 
%$\mathcal{O}$ is the observability matrix, 
%and $\mathcal{B}$ is a combination of the 
%discrete time matrices ($A_d$ and $B_d$) such that it 
%denotes the response of a vector of inputs to states, 
%and $\mathcal{B}_{nN_p-n+1:nN_p}$ denotes the last $n$
%rows of $\mathcal{B}$.

%Since the measurements are a simple linear function 
%of the grey-box and plant states, we use only one 
%neural network to 
%augment the grey-box dynamics. 
%The expected function 
%to be approximated with the network is
%\begin{align*}
%  f_N(\mathbf{y}_{k-N_p:k}, \mathbf{u}_{k+1-N_p:k-1}) \approx 
%  \begin{bmatrix}
%  0 & 0 & 0 \\  
%  0 & -k_2 & k_3 
%  \end{bmatrix}x_p(k)
%\end{align*}

%In this example, the network just approximates a linear function.
%The Figure \ref{fig:validation_linear} 
%below shows validation performances of the grey-box 
%and augmented hybrid models. The structure 
%of the network is restricted to just a matrix and
%the augmented model was trained on 240 samples (4 hours).

%THe Figures \ref{fig:cost_nonlinear_5hours} and 
%\ref{fig:cost_nonlinear_5hours}

%\begin{figure}[!h]
%  \centering
%  \includegraphics[page=2, width=\textwidth]{tworeac_plots_lin.pdf}
%  \caption{Model validation for the linear plant}
%  \label{fig:validation_linear}
%\end{figure}

%We notice that the augmented model accurately predicts the plant 
%behavior. The steady cost profile (Cost = $c_1C_{Af} - c_2C_B$)
%is shown in Figure \ref{fig:cost_linear}.

%\begin{figure}[!h]
%  \centering
%  \includegraphics[page=3, width=\textwidth]{tworeac_plots_lin.pdf}
%  \caption{Steady state cost curve for 
%           the linear plant, grey-box, and hybrid model.}
%  \label{fig:cost_linear}
%\end{figure}

\subsection{Nonlinear reactor example}
We consider a nonlinear reactor with 
two reactions ($A \rightarrow B $ and  $3B \rightleftharpoons C$),
simulated with the following ODEs
\begin{align*}
  \dfrac{dC_A}{dt} &= \dfrac{C_{Af} - C_A}{\tau} - k_1C_A\\
  \dfrac{dC_B}{dt} &= k_1C_A - 3k_2C^3_B + 3k_3C_C- \dfrac{C_B}{\tau}\\
  \dfrac{dC_C}{dt} &= k_2C^3_B - k_3C_C - \dfrac{C_C}{\tau}
\end{align*}
